A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-11-14 Simulation/Writeup/preprint.Rnw
1000 sims results

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data, and ad hoc solutions can yield wildly invalid results \citep{buur18}. To circumvent the ubiquitous problem of missing information, \cite{rubin87} proposed the framework of multiple imputation (MI). MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e. imputed) several times. [Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \cite{rubin87}.
%Missing data is ubiquitous. MICE can solve stuff.

@@

This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \cite{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \cite{mice} in \proglang{R} \cite{R}. It therefore follows notation and conventions of \cite{mice}. For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.

@@

With multiple imputation, $Y_{mis}$ is 'filled in' (i.e. imputed) $m$ times. The $m$ imputed data sets are analyzed separately, and their results pooled. A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputated data sets). The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$. That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values. These are combined into pooled estimate $\bar{Q}$.
% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

@@

Convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains. Ideally, we want the chains to intermingle nicely, without trends within chains. 
% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points. Do these properties hold for the MICE algorithm? Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space. This flexibility is actually the main rationale for FCS instead of a joint model. Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent. A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ . The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 . In general, we would like the statistical inferences to be independent of the stopping point. A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points. The stopping point should not affect the statistical inferences. The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily. Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior. For example, if imputations are made by deterministic functions, the Markov chain may lock up. Such cases can sometimes be diagnosed from the trace lines of the sampler. See Section 6.5.2 for an example. As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent. The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73). The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \cite{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation.

@@

Why is $\widehat{R}$ not applicable on MI data? $\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to. With MI procedures, initial values are chosen such that ... [or maybe even estimated from the observed data??]. This means that at most one initial value is necessary, but probably none at all. [I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \cite{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations. %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50. Therefore, the aim is to replicate empirical finding Lacerda et al. 
Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence. [Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data. We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.

@@

Convergence diagnostics are $\widehat{R}$ and auto-correlation at lag one. $\widehat{R}$ is computed as follows (notation from \cite[p.~5]{veht19}). The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

In this study, $\widehat{R}$ is computed for each variable, in each imputation chain, over all simulation conditions and MCMC simulations. For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.

@@

\begin{Code}
# pseudo-code of simulation 
simulate data with missingness
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    impute the missingness
    compute convergence diagnostic
    perform analysis
    pool results
    compute simulation diagnostics
save convergence and simulation diagnostics
\end{Code}

@@

Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains. The number of iterations is varied over simulations ('maxit' argument between 1 and 100). For each number of iterations, simulation and convergence diagnostics are aggregated over 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

@@

@@

Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?

>>>>

At some point, any scientist conducting statistical analyses will run into a missing data problem \citep{alli02}. Missingness is problematic because statistical inference cannot be performed on incomplete data, and ad hoc solutions can yield wildly invalid results \citep{buur18}. To circumvent the ubiquitous problem of missing information, \cite{rubin87} proposed the framework of multiple imputation (MI). MI is an iterative algorithmic procedure in which missing data points are 'guessed' (i.e. imputed) several times. [Add some stuff here about MICE.] The variability between the imputations validly reflects how much uncertainty in the inference is due to missing information--that is, if all statistical assumptions are met \citep{rubin87}.
%Missing data is ubiquitous. MICE can solve stuff.

@@

This paper investigates convergence properties of MI algorithms by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for applied researchers to evaluate convergence of MI algorithms. %I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. It therefore follows notation and conventions of \cite{mice}. For an overview of the deviations from the 'original' notation by \cite{rubin87}, see \cite{buur18}.

@@

With multiple imputation, $Y_{mis}$ is 'filled in' (i.e. imputed) $m$ times. The $m$ imputed data sets are analyzed separately, and their results pooled. A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputed data sets). The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$. That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values. These are combined into pooled estimate $\bar{Q}$.
% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

[Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), `simulation condition' (the number of iterations the algorithm has before imputing), ...?]

@@

Convergence has two interpretations [look up source!!!]: mixing between chains and stability over iterations within chains. Ideally, we want the chains to intermingle nicely, without trends within chains. 
% "In order to converge to a stationary distribution, a Markov chain needs to satisfy three important properties (Roberts 1996; Tierney 1996): irreducible, the chain must be able to reach all interesting parts of the state space; aperiodic, the chain should not oscillate between different states; recurrence, all interesting parts can be reached infinitely often, at least from almost all starting points. Do these properties hold for the MICE algorithm? Irreducibility is generally not a problem since the user has large control over the interesting parts of the state space. This flexibility is actually the main rationale for FCS instead of a joint model. Periodicity is a potential problem, and can arise in the situation where imputation models are clearly inconsistent. A rather artificial example of an oscillatory behavior occurs when Y 1 is imputed by Y 2 $\beta$ + $\epsilon$  1 and Y 2 is imputed by - Y 1 $\beta$ + $\epsilon$  2 for some fixed, nonzero $\beta$ . The sampler will oscillate between two qualitatively different states, so the correlation between Y 1 and Y 2 after imputing Y 1 will differ from that after imputing Y 2 . In general, we would like the statistical inferences to be independent of the stopping point. A way to diagnose the ping-pong problem, or order effect, is to stop the chain at different points. The stopping point should not affect the statistical inferences. The addition of noise to create imputations is a safeguard against periodicity, and allows the sampler to "break out" more easily. Non-recurrence may also be a potential difficulty, manifesting itself as explosive or non-stationary behavior. For example, if imputations are made by deterministic functions, the Markov chain may lock up. Such cases can sometimes be diagnosed from the trace lines of the sampler. See Section 6.5.2 for an example. As long as the parameters of imputation models are estimated from the data, non-recurrence is mild or absent. The required properties of the MCMC method can be translated into conditions on the eigenvalues of the matrix of transition probabilities (MacKay 2003, 372-73). The development of practical tools that put these conditions to work for multiple imputation is still an ongoing research problem." \cite[par.~4.5]{buur18}
We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation.

@@

Why is $\widehat{R}$ not applicable on MI data? $\widehat{R}$ is not appropriate because it assumes over-dispersed initial values, which means that the initial values of the $m$ imputation chains are 'far away' from the distribution that the chains are converging to. With MI procedures, initial values are chosen such that ... [or maybe even estimated from the observed data??]. This means that at most one initial value is necessary, but probably none at all. [I should look this up.] Without over-dispersed initial states, $\widehat{R}$ can falsely diagnose convergence \citep{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. Empirical finding, however, show that the opposite may be true: \cite{lace07} report $\widehat{R}$ values above the threshold of $\widehat{R} < 1.1$ after fifty iterations. %that $\widehat{R}$ will not be smaller than 1.1 before iteration number 50. Therefore, the aim is to replicate empirical finding Lacerda et al. 
Hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence. [Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data. We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.

@@

Convergence diagnostics are $\widehat{R}$ and auto-correlation (at lag one). $\widehat{R}$ is computed as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

In this study, $\widehat{R}$ is computed for the mean of each variable, in each imputation chain, over all simulation conditions and MCMC simulations. %For each simulation in each simulation condition, the maximum value of $\widehat{R}$ is reported.

@@

\begin{Code}
# pseudo-code of simulation 
simulate data with missingness
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    impute the missingness
    compute convergence diagnostic
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
\end{Code}

@@

Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$). The number of iterations is varied over simulations ('maxit' argument between 1 and 100). Each simulation condition (i.e., each number of iterations) is simulated 1000 times. Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

@@

Post-processing of simulation diagnostics:

Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.

Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%). The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom). CIW is averaged over all simulations with the same simulation condition.

Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.

Post-processing of convergence diagnostics:

$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation. The maximum value across variables within the same simulation is reported. Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.

Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration. For now, we only consider the chain mean as scalar of interest $\theta$. The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

[Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?]

### 2019-11-14 Simulation/Writeup/preprint.tex
1000 sims results

With multiple imputation, $Y_{mis}$ is 'filled in' (i.e. imputed) $m$ times. The $m$ imputed data sets are analyzed separately, and their results pooled. A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputated data sets). The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$. That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values. These are combined into pooled estimate $\bar{Q}$.
% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

@@

>>>>

With multiple imputation, $Y_{mis}$ is 'filled in' (i.e. imputed) $m$ times. The $m$ imputed data sets are analyzed separately, and their results pooled. A distinction is made between the imputation step (employing $p$ 'imputation models'), and the analysis step (using one 'complete-data model' for each of the $m$ imputed data sets). The quantity of scientific interest is the result of the analysis step, e.g., a regression coefficient, and denoted with $Q$. That is, $Q$ is estimated on each imputed data set, resulting in $m$ separate $\hat{Q}$ values. These are combined into pooled estimate $\bar{Q}$.
% "Let Q denote the quantity of scientific interest (e.g., a regression coefficient)." \cite[p.~4]{mice}.

Explain the use of `imputation chain' (the iterations of an imputation, only the final imputed values are used), `simulation condition' (the number of iterations the algorithm has before imputing), ...?

@@

Post-processing of simulation diagnostics:

Bias is computed as the estimated regression coefficient after MI minus the true regression coefficient. Bias is averaged over all simulations with the same simulation condition.
Confidence interval width (CIW) is computed as the diffence between the lower and upper bound of the 95\% confidence interval (CI95\%). The CI95\% bounds are computed as the estimated regression coefficient plus or minus (respectively) the pooled SE across imputations times 2.66 (the quantile of a t distribution with m-1 degrees of freedom). CIW is averaged over all simulations with the same simulation condition.
Coverage rate is computed as the proportion of simulations (with the same simulation condition) in which the true regression coefficient is between the bounds of the CI95\%.

Post-processing of convergence diagnostics:

$\widehat{R}$ is computed within imputation chains: for each variable, for each simulation condition, for each simulation. The maximum value across variables within the same simulation is reported. Maximum $\widehat{R}$ values per simulation condition are aggregated across simulations.

Autocorrelation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration. For now, we only consider the chain mean as scalar of interest $\theta$. The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.
